---
title: "ECON110 Project"
author: "[Student's Name]"
date: "1/26/2022"
output: html_document
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)

# Use the ECON110 Tools file, which is provided by the course's instructor
source("ECON110 Tools.R")

# R libraries are being attached to this project
library(fredr)
library(bea.R)
library(tidycensus)
library(ggplot2)
# library(devtools)
# install_github("mikeasilva/blsAPI")
library(blsAPI)

# My access keys, which will be used in this project
key.FRED   = "3f1722e0293651c883a63c15725200f8"
key.BEA    = "5125F980-B756-4FED-9173-7AADF5DE8231"
key.Census = "302ba19863cd8758a59ade8470d22ca410fd5a05"
key.BLS    = "6b732f6c133d48669aadd2790cb71ccc"

# Registering access keys with data providers
fredr_set_key(key.FRED)
census_api_key(key.Census)

```

## Step No.1

Let's take a look at, for example, the exchange rate history of the US dollar against the euro. FRED collects this data series on a daily basis under ID ```DEXUSEU```. 

The FRED data set has two columns that interest us most: ```date``` and ```value```. 

``` {r echo=FALSE}
# Grab the data on the exchange rate of US dollar to the euro
# Note: the unique ID for this data in the FRED system is DEXUSEU
t <- 
  fredr(
    series_id         = "DEXUSEU",
    observation_start = as.Date("2012-01-01"),
    observation_end   = Sys.Date()
  )

# Let's remove records from t, which does not contain data
t <- na.omit(t)

# Display first 6 records of the data set t
tail(t[,1:3])

```

Now, let's plot the data. See below. 

```{r echo=FALSE}
ggplot(t, aes(x=date, y=value)) + geom_line(col="red")
```

## Non-Farm Payroll

Let's 

  (1) grab the **non-farm payroll** employment data from BLS (ID: ```CES0000000001```), 
  (2) then derive the NFP change in the most recent period and 
  (3) graph the whole data set.

```{r echo=FALSE}

p <- list(                                  # Creating list of data to be retrieved from BLS...
  "seriesid"        = c("CES0000000001"),   # Non-farm payroll employment
  "startyear"       = 2016,                 # ...first year in the time range,
  "endyear"         = year(Sys.Date()),     # ...current year is the end year of the time range 
  "registrationKey" = key.BLS               # BLS key
  )

nfp       <- blsAPI(p,2,TRUE)               # Obtaining the NFP data 
nfp$value <- as.numeric(nfp$value)          # Converting value (thousands of jobs) into numeric
nfp$date  <- as.Date(                       # Creating new variable date to be used in graph 
  paste0("1 ",nfp$periodName," ",nfp$year), 
  format = "%d %B %Y")
nfp <- nfp[order(nfp$date),]                # Ordering data rows by date column
# Most recent NFP increase(+)/decrease(-), unit: thousands of jobs
format(nfp$date[nrow(nfp)],"%B %Y")
diff(nfp$value,1,1)[nrow(nfp)-1]            # The value of the most recent NFP observation

# Plotting the NFP data
ggplot(nfp, aes(x=date,y=value)) + geom_line(col="red")

```

## Consumer Price Index

Let's grab the most recent inflation data from BLS and graph it. I am interested in 
  
  (1) the general consumer price index, which is CPI-**All items** (ID: ```CUUR0000SA0```),  
  (2) the price index of CPI-**Used cars and trucks** (ID: ```CUUR0000SETA02```),
  (3) and the price index of CPI-**Meats** (ID: ```CUUR0000SAF11211```). 

``` {r echo=FALSE}

# Item codes for CPI details: https://download.bls.gov/pub/time.series/cu/cu.item
# Rule for ID construction: CUUR0000 + Item code
# Example 1: item code for "All Items" is SA0
#            hence, ID for CPI in "All items" is CUUR0000SA0
# Example 2: item code for "Used cars and trucks" is SETA02
#            hence, ID for CPI in "Used cars and trucks" is CUUR0000SETA02

p <- list(                                  # Creating list of data to be retrieved from BLS...
  "seriesid"        = c("CUUR0000SA0",      # All items
                        "CUUR0000SETA02",   # Used cars and trucks
                        "CUUR0000SAF11211"),# Meats
  "startyear"       = 2016,                 # Start year in the time range,
  "endyear"         = year(Sys.Date()),     # ...current year is the end year of the time range 
  "registrationKey" = key.BLS               # BLS key
  )              

cpi       <- blsAPI(p,2,TRUE)               # Obtaining the CPI data
cpi$value <- as.numeric(cpi$value)
cpi$date  <- as.Date(paste0("1 ",cpi$periodName," ",cpi$year), format = "%d %B %Y")
cpi       <- cpi[order(cpi$date),]

# Adding new column item_name, containing ID names, into the cpi data set
cpi <- bls.cpi.item_name(cpi, p$seriesid)

# Plotting the CPI data as indexes
ggplot(cpi,     aes(x=date,y=value,color=item_name)) + geom_line()

```

Now, let's derive and plot the year-on-year changes in the CPI indexes:

```{r echo=FALSE}

cpi.yoy <- data.frame()
for (i in 1:length(p$seriesid)) {
    d <- bls.cpi.yoy(cpi,p$seriesid[i])
    cpi.yoy   <- rbind(cpi.yoy, d)
}

# Plotting the CPI data as year-on-year changes
ggplot(cpi.yoy, aes(x=date,y=value,color=item_name)) + geom_line()

```
