---
title: "ECON110 Project"
author: "Oleksandr Valchyshen"
date: "1/26/2022"
output: html_document
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)

# Use the ECON110 Tools file, which is provided by the course's instructor
source("ECON110-Tools.R")

# R libraries are being attached to this project
library(fredr)
library(bea.R)
library(tidycensus)
library(ggplot2)
library(blsAPI)

# My access keys, which will be used in this project
key.FRED   = "3f1722e0293651c883a63c15725200f8"
key.BEA    = "5125F980-B756-4FED-9173-7AADF5DE8231"
key.Census = "302ba19863cd8758a59ade8470d22ca410fd5a05"
key.BLS    = "6b732f6c133d48669aadd2790cb71ccc"

# Registering access keys with data providers
fredr_set_key(key.FRED)
census_api_key(key.Census)

```

## Step No.1

Let's take a look at, for example, the exchange rate history of the US dollar against the euro. FRED collects this data series on a daily basis under ID ```DEXUSEU```. 

The FRED data set has two columns that interest us most: ```date``` and ```value```. 

``` {r echo=FALSE}
# Grab the data on the exchange rate of US dollar to the euro
# Note: the unique ID for this data in the FRED system is DEXUSEU
t <- 
  fredr(
    series_id         = "DEXUSEU",
    observation_start = as.Date("2012-01-01"),
    observation_end   = Sys.Date()
  )

# Let's remove records from t, which does not contain data
t <- na.omit(t)

# Display first 6 records of the data set t
tail(t[,1:3])

```

Now, let's plot the data. See below. 

```{r echo=FALSE}

ggplot(t, aes(x=date, y=value)) + geom_line(col="red")

```

## Non-Farm Payroll

Let's 

  (1) grab the **non-farm payroll** employment data from BLS (ID: ```CES0000000001```), 
  (2) then derive the NFP change in the most recent period and 
  (3) graph the whole data set.

```{r echo=FALSE}

p <- list(                                  # Creating list of data to be retrieved from BLS...
  "seriesid"        = c("CES0000000001"),   # Non-farm payroll employment
  "startyear"       = 2016,                 # ...first year in the time range,
  "endyear"         = year(Sys.Date()),     # ...current year is the end year of the time range 
  "registrationKey" = key.BLS               # BLS key
  )

nfp       <- blsAPI(p,2,TRUE)               # Obtaining the NFP data 
nfp$value <- as.numeric(nfp$value)          # Converting value (thousands of jobs) into numeric
nfp$date  <- as.Date(                       # Creating new variable date to be used in graph 
  paste0("1 ",nfp$periodName," ",nfp$year), 
  format = "%d %B %Y")
nfp <- nfp[order(nfp$date),]                # Ordering data rows by date column
# Most recent NFP increase(+)/decrease(-), unit: thousands of jobs
format(nfp$date[nrow(nfp)],"%B %Y")
diff(nfp$value,1,1)[nrow(nfp)-1]            # The value of the most recent NFP observation

# Plotting the NFP data
ggplot(nfp, aes(x=date,y=value)) + geom_line(col="red")

```

## Consumer Price Index

Let's grab the most recent inflation data from BLS and graph it. I am interested in 
  
  (1) the general consumer price index, which is CPI-**All items** (ID: ```CUUR0000SA0```),  
  (2) the price index of CPI-**Used cars and trucks** (ID: ```CUUR0000SETA02```),
  (3) and the price index of CPI-**Meats** (ID: ```CUUR0000SAF11211```). 

``` {r echo=FALSE}

# Item codes for CPI details: https://download.bls.gov/pub/time.series/cu/cu.item
# Rule for ID construction: CUUR0000 + Item code
# Example 1: item code for "All Items" is SA0
#            hence, ID for CPI in "All items" is CUUR0000SA0
# Example 2: item code for "Used cars and trucks" is SETA02
#            hence, ID for CPI in "Used cars and trucks" is CUUR0000SETA02

p <- list(                                  # Creating list of data to be retrieved from BLS...
  "seriesid"        = c("CUUR0000SA0",      # All items
                        "CUUR0000SETA02",     # Used cars and trucks
                        "CUUR0000SAF11211"),  # Meats
  "startyear"       = 2016,                 # Start year in the time range,
  "endyear"         = year(Sys.Date()),     # ...current year is the end year of the time range 
  "registrationKey" = key.BLS               # BLS key
  )              

cpi       <- blsAPI(p,2,TRUE)               # Obtaining the CPI data
cpi$value <- as.numeric(cpi$value)
cpi$date  <- as.Date(paste0("1 ",cpi$periodName," ",cpi$year), format = "%d %B %Y")
cpi       <- cpi[order(cpi$date),]

# Adding new column item_name, containing ID names, into the cpi data set
cpi <- bls.cpi.item_name(cpi, p$seriesid)

# Plotting the CPI data as indexes
ggplot(cpi,     aes(x=date,y=value,color=item_name)) + geom_line()

```

Now, let's derive and plot the year-on-year changes in the CPI indexes:

```{r echo=FALSE}

cpi.yoy <- data.frame()
for (i in 1:length(p$seriesid)) {
    d <- bls.cpi.yoy(cpi,p$seriesid[i])
    cpi.yoy   <- rbind(cpi.yoy, d)
}

# Plotting the CPI data as year-on-year changes
ggplot(cpi.yoy, aes(x=date,y=value,color=item_name)) + geom_line()

```

## Norway

Money of account is Norwegian krone.

Central Bank of Norway: https://www.norges-bank.no/en/

Ministry of Finance: https://www.regjeringen.no/en/dep/fin/id216/

### GDP (ID: CLVMNACSCAB1GQNO)

``` {r echo=FALSE}
# Grab the data on the exchange rate of US dollar to the euro
# Note: the unique ID for this data in the FRED system is DEXUSEU
t <- 
  fredr(
    series_id         = "CLVMNACSCAB1GQNO",
    observation_start = as.Date("1950-01-01"),
    observation_end   = Sys.Date()
  )

# Let's remove records from t, which does not contain data
t <- na.omit(t)

# Display first 6 records of the data set t
tail(t[,1:3])

```

Now, let's plot the data. See below. 

```{r echo=FALSE}

ggplot(t, aes(x=date, y=value)) + geom_line(col="red")

# GDP in real and seasonally-adjusted terms as a time series
y <- as.numeric(format(t$date[1],"%Y"))
q <- as.numeric(substr(quarters(t$date[1]),2,2))
gdp.ts <- ts(data = t$value, start = c(y,q), frequency = 4)

```

__Determining the most recent business cycle__

First step is to observe the chart above and spot two most recent peaks on the chart. Decide approximately after which year did the left-side peak take place. In the case of Norway it is 2015.

Second step is to narrow the time series data on the country's GDP ```gdp.ts```. And then to build historical tables of (1) values of GDP in "chained" domestic money of account, and (2) real GDP growth rates in % SAAR terms. 

```{r echo=TRUE}

# Narrowing down the times series of the country's GDP adjusted for inflation
gdp.ts <- window(gdp.ts, start = c(2015-1,1), frequency = 4)

# GDP values in "chained" local money of account 
#(adjusted for inflation and seasonality)
gdp.ts

# Real GDP growth rate, % SAAR
((gdp.ts/stats::lag(gdp.ts,-1))^4-1)*100 

# Plotting the data set with country's GDP values to spot the peaks
plot(gdp.ts, 
     main = "Quarterly GDP, showing most recent business cycle",
     sub  = "Millions of Norwegian krone, adjusted for inflation and seasonality",
     xlab = "", 
     ylab = "")

```

So, Peak #1 is 3Q (October) of 2015 and Peak #2 is 4Q (December) of 2019. Hence, __length of the business cycle is number of months between Oct-2015 and Dec-2019, which is 51 months (see below: 51 = 50 + 1) or 4 years and 1 quarter__. 

```{r echo=TRUE}

# By using R function called interval(), we can obtain the number of 
# months between two dates
interval(start = as.Date("2015-10-01"), end = as.Date("2019-12-01")) %/% months(1)

```

### Inflation (ID: NORCPIALLMINMEI)

``` {r echo=FALSE}
# Grab the data for Norway's inflation
t <- 
  fredr(
    series_id         = "NORCPIALLMINMEI",
    observation_start = as.Date("2012-01-01"),
    observation_end   = Sys.Date()
  )

# Let's remove records from t, which does not contain data
t <- na.omit(t)

# Display first 6 records of the data set t
tail(t[,1:3])

```

Now, let's plot the data. See below. 

```{r echo=FALSE}

ggplot(t, aes(x=date, y=value)) + geom_line(col="red")

# CPI index as a time series
y <- as.numeric(format(t$date[1],"%Y"))
m <- as.numeric(month(t$date[1]))
cpi.ts <- ts(data = t$value, start = c(y,m), frequency = 12)
# CPI as year-on-year time series
cpi.yoy.ts <- (cpi.ts/stats::lag(cpi.ts,-12)-1)*100 

# CPI time series plot 
plot(cpi.yoy.ts,
     main = "History of Norway's CPI (% year-on-year)",
     xlab = "",
     ylab = "")

window(cpi.yoy.ts, start=c(2020,1))

```

## Interest Rates: USA

The corridor system of interest rates is the range of interest rates the central bank maintains daily in order to keep the overnight interest rate within the targeted range (corridor). 

```{r echo=FALSE} 

ir.corridor.us("1990-01-01")
#ir.corridor.ca("1990-01-01")
#ir.corridor.mx("1990-01-01")
#ir.corridor.tr("1990-01-01")

```

Government bonds of different maturities have yields that fluctuate depending on the level of the overnight interest rate that follows the target set by the central bank's policy.

```{r echo=FALSE} 

ir.govt.us("1990-01-01")
#ir.govt.ca("1990-01-01")
#ir.govt.mx("1990-01-01")
#ir.govt.tr("1990-01-01")

```
